Improving Item-Item Similarity Estimation during Collaborative Filtering

### 2014-03-25 paper/paper.tex
Add a couple of paragraphs to Model-based Technique section.

\section*{Introduction}
Determining the similarity of two users or items is valuable for a
number of data-mining applications including recommender systems, product
assortment and differentiation, as well as branding. Collaborative filtering is
often used to exploit user-item rating data to make such estimates. This is 
typically done by computing the Pearson correlation or cosine similarity of 
vector representations of the users or items \cite{Su2009}. Given sufficient 
user-item rating data, this approach can be very effective. However, when the
rating data is sparse for one or both users/items, these metrics can
yield dramatic discrepancies compared to values computed with sufficient
data at a later point in time. In this report, we describe two novel techniques 
for estimating the true similarity of two users or items given limited common 
rating data.

To keep the present work focused, we confine ourselves to memory-based
collaborative filtering (CF) in which we work with a logical user-item matrix.
Each element of the matrix corresponds to a rating given by a particular user to a
particular item. Further, we choose to limit our analysis to item-item
similarity but note that these techniques should be equally effective at
improving user-user similarity estimation. In this representaion, an item is
a vector of the ratings it has received from all users. We now formalize the 
notion of item-item similarity. Two alternative models for the generation of 
user-item ratings are useful in arriving at two different definitions of the 
{\em true} item-item similarity.

@@

.....

Even worse, low confidence measurements of similarity are indistinguishable from
high confidence ones. In this paper, we develop and compare a number of methods
to produce more accurate estimates of similarity by incorporating confidence
into the similarity score.

Item-based collaborative filtering is a common technique for measuring the
similarity of two items. There are different forms of collaborative filtering.
For this paper, we use a traditional approach. Each item is represented as a
vector of ratings. The similarity score of two items is computed by measuring
the similarity of the two rating vectors.

There are a few ways to compute the similarity of two vectors. One approach is
to use Cosine-similarity, defined as the cosine of the angle between the two
vectors:
\begin{align}
CosSim(A, B) = \frac{\sum\limits_{u\in U_{AB}}
r_{u,A}r_{u,B}}{\sqrt{\sum\limits_{u\in U_{A}} r_{u,A}^2}
\sqrt{\sum\limits_{u\in U_{B}} r_{u,B}^2}}
\end{align}
where $U_{I}$ is the set of all users who rated item $I$, $U_{AB}$ is the set of
all users who rated both item $A$ and item $B$ and $r_{u,I}$ is the rating user
$u$ gave to item $I$. Another popular approach is to use Pearson correlation:
\begin{align}
PearsSim(A, B) = \frac{\sum\limits_{u\in U_{AB}}
r_{u,A}r_{u,B}}{\sqrt{\sum\limits_{u\in U_{AB}} r_{u,A}^2}
\sqrt{\sum\limits_{u\in U_{AB}} r_{u,B}^2}}
\end{align}

Other similarity functions exist, such as one-sided similarities, but Pearson
correlation and Cosine-similarity are the most popular and will be the two
measurements used in this paper. \footnotemark

\footnotetext{ Note that the difference between the two similarities is how they
handle unpaired ratings; that is ratings from a user who has not rated the other
item. Cosine-similarity considers the unknown rating from an unpaired rating to
be $0$ and then calculates the cosine of the two vectors. Pearson correlation
simply throws away all unpaired ratings and calculates the cosine of the two
modified vectors.}

Both similarity measurements primarily leverage information from common users,
that is users who have rated both items. Because of this, they perform well with
a large number common users, but give unreliable results when the items have few
common users. In an extreme case, if only one user has rated item $A$ and item
$B$, $PearsSim(A, B) = 1$ or $-1$. Not only is this unlikely to be an accurate
assessment of the true similarity of $A$ and $B$, it also gives the most extreme
results possible, without giving any indication that this is a low-confidence
calculation.

In this paper, we construct a more accurate similarity score by incorporating
the number of common users, $n$, into the similarity function.

@@

>>>>

\section*{Introduction}
Determining the similarity of two users or items is valuable for a number of
data-mining applications including recommender systems, product assortment and
differentiation, as well as branding. Collaborative filtering is often used to
exploit user-item rating data to make such estimates. This is typically done by
computing the Pearson correlation or cosine similarity of vector representations
of the users or items \cite{Su2009}. Given sufficient user-item rating data,
this approach can be very effective. However, when the rating data is sparse for
one or both users/items, these metrics can yield dramatic discrepancies compared
to values computed with sufficient data at a later point in time. Moreover,
these early, low-confidence measurements can be indistinguishable from later,
high-confidence measurements.  In this report, we describe two novel techniques
for estimating the true similarity of two users or items given limited common
rating data by incorporating confidence into the measurement.

To keep the present work focused, we confine ourselves to memory-based
collaborative filtering (CF) in which we work with a logical user-item matrix.
Each element of the matrix corresponds to a rating given by a particular user to
a particular item. Further, we choose to limit our analysis to item-item
similarity but note that these techniques should be equally effective at
improving user-user similarity estimation. In this representation, an item is a
vector of the ratings it has received from all users. We now formalize the
notion of item-item similarity. Two alternative models for the generation of
user-item ratings are useful in arriving at two different definitions of the
{\em true} item-item similarity.

@@

Both similarity estimates primarily leverage information from common users, that
is users who have rated both items. Because of this, they perform well with a
large number common users, but give unreliable, and dramatically varying,
results when the items have few common users. In an extreme case, if only one
user has rated item $A$ and item $B$, $PearsSim(A, B) = 1$ or $-1$. Not only is
this unlikely to be an accurate assessment of the true similarity of $A$ and
$B$, it also gives the most extreme results possible, without giving any
indication that this is a low-confidence calculation.

@@

\section*{A Model-based Technique}

We now construct a relatively simple probabilistic model that is readily
exploited for making estimates of $TrueSimRand$ and, to a lesser extent,
$TrueSimPref$. In addition, this approach has the added benefit of allowing us
to infer implied predictions of $TrueSimRand$ made by individual users each time
they rate two items. This latter feature allows us to define a new metric for
users that we call {\em user-predictivity}, which allows us to ascribe varying
levels of confidence to their contributions to a similarity estimate.

Let $Y$ be a random variable representing the $TrueSimRand$ of a randomly chosen
pair of items. As before, we model $Y$ as a Normal distribution
\begin{align}
Y \sim N(\mu_s, \sigma_s^2),
\end{align}
where $\mu_s$ and $\sigma_s$ are the average and standard deviation of all
similarity scores, respectively. Now, let $\vec{R} = \left(R_A~~R_B\right)$ be
a random two element vector representing the ratings given to items $A$ and $B$.
We model the conditional distribution of $\vec{R} | Y$ as a multivariate Normal
distribution
\begin{align}
\vec{R} | Y=y \sim N(\vec{\mu_r}, \mathbf{\Sigma_y}),
\end{align}
where the elements of $\vec{\mu_r}$ are the average ratings and
$\mathbf{\Sigma_y}$ is the $y$-dependent $2\times2$ covariance matrix.
